Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-10 paper/introduction.tex
updates to intro

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of potentially more than one contiguous
range of memory locations, permitting a region to grow) that is used to realize
a data-structure with a well-defined lifetime. In particular, the entire region
is typically freed in one shot.

Region-based memory management, both manual as well as automatic, has been known for
a long time. Manual region-based memory management suffers from the usual challenges,
namely the potential for dangling references and lack of memory safety. Automatic
region-based memory managements guarantee memory safety, but impose various restrictions.
Tofte and Talpin's system, for example, requires that the set of all regions (existing
at a point in time) forms a stack. Specifically, this requires that the lifetimes of
all regions must be well-nested: it is not possible to have two regions whose lifetimes
overlap, with neither one's lifetime contained within the other.
Unfortunately, the data-structures in the above example do not satisfy this restriction.
We refer to such regions as \emph{dynamic} regions.

The goal of this work is an automatic (memory-safe) region-based memory management technique
that supports dynamic regions (that are first-class objects) and higher-order functions.
Though our focus is on memory-safety, dynamic regions also have other side-benefits.
When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
provides an ownership discipline that ensures that the producer cannot subsequently modify
the data-structure. When the same transfer happens in a distributed setting, the system guarantees
that the data-structure is self-contained and can be safely transmitted across address spaces.
\TODO{How accurate is this claim?}

The key memory safety property we wish to ensure is that there are no dangling references:
i.e., a reference to an object that has been freed.
\TODO{Or, should we say: no dangling reference is dereferenced}
A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
(In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
when a garbage collector promotes an object from a lower generation to a higher generation.
From this perspective, a region may be seen as a user-controlled generation.)

@@

We combine \emph{static} regions with the following protocol to ensure safety in these cases.
To work with a dynamic region $D$, we must first \emph{open} region D1,
to indicate that $D$ should not be freed during this period.
When the processing of the region is complete, we close the region $D$,
to indicate that it is safe to free $D$.
Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
guaranteed to be contained within the interval when $D$ is open.
As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
no pointer from outside $D$ points to an internal object of $D$.

The final piece required to ensure memory-safety is ensuring that the program correctly follows the above-mentioned
protocol for dynamic regions. The protocol is essentially a typestate specification: an open region cannot be freed and,
dually, a freed region cannot be opened.

Dynamic regions are first-class objects in our system: e.g., in our running example,
we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
aliases (pointers) to the same region, which, in turn, means that statically checking if the program correctly follows the
dynamic region protocol is hard (undecidable, in fact). Safety can be ensured by using, \eg, linear typing to prevent the
creation of aliases, but this would be quite restrictive. Hence, in our system, we dynamically check for adherence to
this protocol.

In summary, our approach reduces memory safety to two sub-problems:
(a) Ensuring  low-level invariants about pointers to (fine-grained) objects inside regions and 
(b) Ensuring higher-level (typestate) invariants about the regions themselves. 
We use a type system to statically ensure the first set of invariants. We use a dynamic check to ensure that the
second set of invariants are satisfied.
%
This is a novel aspect of our system. We believe that this is a reasonable choice:
dynamically checking for the second category of invariants is easy and does not pose a performance concern (unlike, \eg, the first category of constraints);
a static typechecker that rules out the second category of violations would be very restrictive from an expressiveness perspective;
finally, the human effort required to ensure the second set of invariants is not as burdensome (as regions are coarse-grained objects
manipulated infrequently), \eg, as it would be for the first set of invariants.

>>>>

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow) that is used to realize
a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is typically freed in one shot.

Region-based memory management, both manual as well as automatic, has been known for
a long time. Manual region-based memory management suffers from the usual challenges,
namely the potential for dangling references and lack of memory safety. Automatic
region-based memory management systems guarantee memory safety, but impose various restrictions.
Tofte and Talpin's system, for example, requires that the set of all regions (existing
at a point in time) forms a stack. Specifically, this requires that the lifetimes of
all regions must be well-nested: it is not possible to have two regions whose lifetimes
overlap, with neither one's lifetime contained within the other.
Unfortunately, the data-structures in the above example do not satisfy this restriction.
We refer to such regions as \emph{dynamic} regions.

The goal of this work is an automatic (memory-safe) region-based memory management technique
that supports dynamic regions (that are first-class objects) and higher-order functions.
Though our focus is on memory-safety, dynamic regions also have other side-benefits.
When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
provides an ownership discipline that ensures that the producer cannot subsequently modify
the data-structure. When the same transfer happens in a distributed setting, the system guarantees
that the data-structure is self-contained and can be safely transmitted across address spaces.

The key memory safety property we wish to ensure is that there are no dangling references:
i.e., a reference to an object that has been freed.
A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
(In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
when a garbage collector promotes an object from a lower generation to a higher generation.
From this perspective, a region may be seen as a user-controlled generation.)

@@

We use \emph{static regions} (regions with lexically-scoped lifetimes), as well as the stack,
for such temporary pointers (to objects inside a dynamic region) and use the following protocol
to ensure safety of such pointers.
To work with a dynamic region $D$, we must first \emph{open} region $D$,
to indicate that $D$ should not be freed during this period.
When the processing of the region is complete, we close the region $D$,
to indicate that it is safe to free $D$.
Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
guaranteed to be contained within the interval when $D$ is open.
As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
no pointer from outside $D$ points to an internal object of $D$.

The final piece required to ensure memory-safety is ensuring that the program correctly follows the above-mentioned
protocol for dynamic regions: \eg, an open region should not be freed and, dually, a freed region should not be opened.
(A complete typestate specification of the protocol is presented later.)

Dynamic regions are first-class objects in our system: \eg, in our running example,
we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
aliases (pointers) to the same region, which, in turn, means that statically checking if the program correctly follows the
dynamic region protocol is hard (undecidable, in fact). Safety can be ensured by using, \eg, linear typing to prevent the
creation of aliases, but this would be quite restrictive. Hence, in our system, we dynamically check for adherence to
this protocol.

In summary, our approach reduces memory safety to two sub-problems:
(a) Ensuring  low-level invariants about pointers to (fine-grained) objects inside regions and 
(b) Ensuring higher-level (typestate) invariants about the regions themselves. 
We use a type system to statically ensure the first set of invariants. We use a dynamic check to ensure that the
second set of invariants are satisfied at runtime.
%
This is a novel aspect of our system. We believe that this is a reasonable choice:
dynamically checking for the second category of invariants is easy and does not pose a performance concern (unlike, \eg, the first category of constraints);
a static typechecker that rules out the second category of violations would be very restrictive from an expressiveness perspective;
finally, the human effort required to ensure the second set of invariants is not as burdensome (as regions are coarse-grained objects
manipulated infrequently), \eg, as it would be for the first set of invariants.
